High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-11-07 PNAS_manuscript/PNAS_manuscript.tex

% Please add a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors leading to its failure in providing reliable paleointensity estimates, the anorthosite xenoliths often yield high-quality, single-slope paleointensity results that are consistent at specimen and site levels. The high paleointensity estimates from the anorthosites require an existence of a vigorous late Mesoproterozoic geodynamo. Together with previous paleointensity data from the Midcontinent Rift, our new results suggest that a strong geodynamo likely persisted through $\sim$25 Myr at ca. 1.1 Ga, inconsistent with the interpretation of a progressive decay of Earth's geodynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$ 3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first-principles calculations implied higher conductivities \cite{Koker2012a,Pozzo2012a} and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a, Zhang2020b}. With the estimates for the core thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of the core and mantle.

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence of an existence of an appreciable geomagnetic field prior to the Ediacaran (Figure \ref{fig:PINT_compilation}). On the other hand, studies have found very low virtual dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \citeA{Bono2019a} adopted a billion-year-scale view of the Precambrian paleointensity records and based on a polynomial fit through selected data to suggest that the geodynamo power progressively decreased through the Proterozoic (Figure \ref{fig:PINT_compilation}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting the geodynamo to have reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Figure \ref{fig:PINT_compilation}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

@@

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \citeA{Vervoort2007a, Miller2013a}) was punctuated by rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \citeA{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Figure \ref{fig:Geologic_map}) formed as a result of having upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's North Eastern Minnesota (Figure \ref{fig:Geologic_map}; \citeA{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Figure \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from \citeA{Pesonen1983a} are categorized by lithology. All sites from previous studies are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of magmatism during the Midcontinent Rift main stage activity. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constraints the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters \cite{Boerboom2004a, Boerboom2006b}. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Figure \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Figure \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

\begin{SCfigure*}[\sidecaptionrelwidth][t]
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \citeA{Zhang2021b}). Petrographic images (A-D) show Duluth Complex anorthosite have euhedral, elongate plagioclase crystals with interlocking texture and igneous foliation. (C) The plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic dipole fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{SCfigure*}

Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \citeA{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \citeA{Haggerty1976a} in other slowly cooled, mafic intrusions. \citeA{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Figure \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Figure \ref{Petro_QDM}).

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots (Arai plots; 
\cite{Arai1963a}) are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table \ref{tab:PINT_result} and are plotted in Figure \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments displayed on Arai and Zijderveld plots (insets) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent our fitting for the dominant single-slope component that passes our acceptance criteria with estimates of the ancient field (B$_{anc}$) being used to determine site means. The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Plot for diabase specimen BD2-5a shows typical paleointensity behavior of this site - specimens typically have double-slope behaviors although the lower temperature slopes can pass the selection criteria, resulting in the high estimated B$_{anc}$ values shown in Figure \ref{fig:PINT_cooling_corrected}. Thus, the estimates of B$_{anc}$ for this specimen plot is illustrative, but were not used to calculate a site mean paleointensity value in Figure \ref{fig:PINT_cooling_corrected}.}}
\label{fig:IZZI_examples}
\end{figure*}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, such records are often difficult to recover due to the alterations associated with weathering after the formation of rocks. In this study, we use plagioclase cumulates from the North American Midcontinent Rift that has experienced minimal post-formation alterations to obtain a record of $\sim$89 ZAm$^2$ virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors leading to its failure in providing reliable paleointensity estimates, the anorthosite xenoliths often yield high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment estimates from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong dipole strength recorded by the anorthosites require an existence of a vigorous late Mesoproterozoic geodynamo. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results suggest that a dynamo with strong power sources likely persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first-principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Labrosse2015a, Ohta2016a, Zhang2020b}. With the estimates for the core thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of the core and mantle.

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence that an appreciable geomagnetic field existed prior to the Ediacaran (Fig. \ref{fig:PINT_compilation}). On the other hand, studies have found very low virtual dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \cite{Bono2019a} adopted a billion-year-scale view of the Precambrian paleointensity records and based on a polynomial fit through selected data to suggest that the geodynamo power progressively decreased through the Proterozoic (Fig. \ref{fig:PINT_compilation}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting the geodynamo to have reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rocks in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual dipole moment (VDM) values similar to that of today, which are higher than the prediction of the long term decay curve (Fig. \ref{fig:PINT_compilation}). \cite{Pesonen1983b} obtained an average virtual dipole moment of 93 $\times 10^21$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). \cite{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \cite{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \cite{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \cite{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \cite{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

@@

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \cite{Vervoort2007a, Miller2013a}) was punctuated by rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \cite{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Fig. \ref{fig:Geologic_map}) formed as a result of upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's northeastern Minnesota (Fig. \ref{fig:Geologic_map}) \cite{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Fig. \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from \cite{Pesonen1983a} are categorized by lithology. All sites from \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of magmatism during the Midcontinent Rift main stage activity. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constrain the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters \cite{Boerboom2004a, Boerboom2006b}. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

@@

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \cite{Zhang2021b}). Petrographic image (A) shows Duluth Complex anorthosite have euhedral, interlocking plagioclase crystals showing igneous foliation. (B) Beaver River anorthosite have plagioclase with a granoblastic texture. (C) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) at UC Berkeley Paleomagnetism Lab show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}

Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \cite{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \cite{Haggerty1976a} in other slowly cooled, mafic intrusions. \cite{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots 
\cite{Arai1963a} are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table \ref{tab:PINT_result} and are plotted in Fig. \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments displayed on Arai and Zijderveld plots (insets) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent our fitting for the dominant single-slope component that passes our acceptance criteria with estimates of the ancient field (B$_{anc}$) being used to determine site means. The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Plot for diabase specimen BD2-5a shows typical paleointensity behavior of this site - specimens typically have double-slope behaviors although the lower temperature slopes can pass the selection criteria, resulting in the high estimated B$_{anc}$ values shown in Fig. \ref{fig:PINT_cooling_corrected}. Thus, the estimates of B$_{anc}$ for this specimen plot is illustrative, but were not used to calculate a site mean paleointensity value in Fig. \ref{fig:PINT_cooling_corrected}.}}
\label{fig:IZZI_examples}
\end{figure*}

The Beaver River anorthosite specimens yielded a high pass rate ($\sim$46.5\%) through our selection criteria. In distinct contrast, the Beaver River diabase specimens are all rejected. Paleointensity experiment results from both lithologies show three types of dominant specimen behaviors on Arai plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope, and zigzagging or sagging. Note that almost all specimens have minimal secondary remanence component in thermal demagnetization results as shown in example orthogonal plots of all zero-field steps (Fig. \ref{fig:IZZI_examples}). This reflects that the rocks have experienced mild alteration after their formation. Typical specimen paleointensity behavior of the anorthosites display straight, single-slope Arai plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity estimates from the anorthosite xenoliths that pass the selection. All diabase specimens failed the selection because of double-slope behavior (fail FRAC selection; see methods), poor pTRM checks, and zigzagging (fail SCAT, DANG selection; see methods) Arai plots (Fig. \ref{fig:IZZI_examples}). As is shown in Fig. \ref{fig:IZZI_examples}, specimen BD2-5b has a high-temperature slope with zigzagging behavior, poor pTRM checks, and low NRM/TRM ratio, in addition to a low-temperature slope with a steep NRM/TRM ratio. Although the BD2 specimens' lower temperature fraction of the Arai plots are straight and the pTRM checks are consistent, lower temperature slopes from BD2 give consistently higher paleointensity estimates than those from site AX6, AX8, and AX10, which are anorthosite xenoliths included within BD2 (Fig. \ref{fig:Geologic_map}) \cite{Zhang2021a}. This higher estimate from the diabase lithology could be attributed to the systematic bias of using the lower temperature slope of the Arai plots demonstrated in Fig. \ref{fig:IZZI_examples} and specimens from site BD2 all fail the criteria of having a maximum magnetization gap between adjacent steps of 25\% of the total NRM. Eventually, we only accept the single-slope anorthosite specimens results for further analyses.
